---
title: "Plastic Pirates EU Analsysis"
output: html_notebook
---

# Paper's Skeleton
The idea is to have two papers
Questions to be solved

* Calification/descualification rate
  * We need to analyse the raw data in order to have those sampling events that were qualified and which doesn't (RiversideWasteVariety_AttemptedSampling == 1 and RiversideWasteVariety_Qualified == 1)
  

# Litter Presence/Absence

# Community-based approach to analyze composition of litter

# Look for explanatory variables inside the dataset from Plastic Pirates. (se está dejando para después)

# Include extra-explanatory variables for all the variables will be nice if something else can take care on this.

# Next meeting before the Workshop on Data Validation/Consolidation and Analysis.
## Focused in the general aspect of data (Types, Kg, Zones, etc.)

# We have to include the informaton of group D (look at quantitative/semi-quantitative data)

# Data Analysis

## Loading Packages

Loading R packages necessary to perform the __EDA__.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(writexl)
library(DT)
library(lubridate)
```

## Data loading

```{r DataLoading, message=FALSE, warning=FALSE}
# Cargamos los datos incorporando las modificaciones para trabajar con él
# Cuidado con el nombre de los ríos, hay besòs y bessos.
Riberbank <- 
  read.csv(file = "PlasticPirates_Riverbank_data/Riverbank_data_v2.0.0.csv", header = TRUE, sep = ",", na.strings = "na") %>% 
  mutate(SamplingDate = lubridate::ymd(SamplingDate)) %>% 
  mutate(across(where(is.character), as.factor))
```

## RESEARCH QUESTIONS

# Citizens Description?

```{r EDACountries, message=FALSE, warning=FALSE}
Riberbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n())
```
```{r EDACountriesPlot, message=FALSE, warning=FALSE}
Riberbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n(), .groups = 'drop') %>%
    # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Num_Sampling = 0)) %>%
  # Ordenar por total acumulado
  mutate(CountryResponsible = reorder(CountryResponsible, -Num_Sampling, sum)) %>%
  ggplot(aes(x = CountryResponsible, y = Num_Sampling, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "dodge", colour = 'black') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#8DEEEE", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of samplings campaigns",
       fill = "Sampling Year")
```


```{r}
Riberbank %>% 
  #select(SamplingRiver) %>% table() 
  group_by(CountryResponsible) %>% 
  summarise(Rivers_total = n_distinct(SamplingRiver))
```

