---
title: "Plastic Pirates EU Analsysis"
output: html_notebook
---

# Paper's Skeleton
The idea is to have two papers
Questions to be solved

* Calification/descualification rate
  * We need to analyse the raw data in order to have those sampling events that were qualified and which doesn't (RiversideWasteVariety_AttemptedSampling == 1 and RiversideWasteVariety_Qualified == 1)
  

# Litter Presence/Absence

# Community-based approach to analyze composition of litter

# Look for explanatory variables inside the dataset from Plastic Pirates. (se está dejando para después)

# Include extra-explanatory variables for all the variables will be nice if something else can take care on this.

# Next meeting before the Workshop on Data Validation/Consolidation and Analysis.
## Focused in the general aspect of data (Types, Kg, Zones, etc.)

# We have to include the informaton of group D (look at quantitative/semi-quantitative data)

# Data Analysis

## Loading Packages

Loading R packages necessary to perform the __EDA__.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(writexl)
library(DT)
library(lubridate)
library(scales)
```

## Data loading

```{r DataLoading, message=FALSE, warning=FALSE}
# Cargamos los datos incorporando las modificaciones para trabajar con él
# Cuidado con el nombre de los ríos, hay besòs y bessos.
Riverbank <- 
  read.csv(file = "PlasticPirates_Riverbank_data/Riverbank_data_v2.0.0.csv", header = TRUE, sep = ",", na.strings = "na") %>% 
  mutate(SamplingDate = lubridate::ymd(SamplingDate)) %>% 
  mutate(across(where(is.character), as.factor))
```

## RESEARCH QUESTIONS

### Citizens Description

#### Numer of samplings per year and country

* Table 

```{r EDACountries, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n())
```
* Plot

```{r EDACountriesPlot, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n(), .groups = 'drop') %>%
    # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Num_Sampling = 0)) %>%
  # Ordenar por total acumulado
  mutate(CountryResponsible = reorder(CountryResponsible, -Num_Sampling, sum)) %>%
  ggplot(aes(x = CountryResponsible, y = Num_Sampling, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "dodge", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#E0FFFF", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of samplings campaigns",
       fill = "Sampling Year")
```

#### Numer of Rivers sampled per year and country

*  Table

```{r EDARivers, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver))
```
* Plot (without numbers)

```{r EDARiversPlot, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver), .groups = 'drop') %>%
  # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Rivers = 0)) %>%
  # Ordenar por total acumulado de ríos
  mutate(CountryResponsible = reorder(CountryResponsible, -Rivers, sum)) %>%
  ggplot(aes(x = CountryResponsible, y = Rivers, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "stack", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#E0FFFF", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of rivers",
       fill = "Sampling Year")
```

* Plot (with numbers)

```{r EDARiversPlotNum, message=FALSE, warning=FALSE}
# Preparar los datos
plot_data <- Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver), .groups = 'drop') %>%
  # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Rivers = 0)) %>%
  # Ordenar por total acumulado de ríos
  mutate(CountryResponsible = reorder(CountryResponsible, -Rivers, sum))

# Crear el gráfico
ggplot(plot_data, aes(x = CountryResponsible, y = Rivers, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "stack", colour = 'black') +
  # Añadir etiquetas solo para valores superiores a 20
  geom_text(data = filter(plot_data, Rivers > 8), 
            aes(label = Rivers), 
            position = position_stack(vjust = 0.5),
            color = "white", fontweight = "bold", size = 2.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#E0FFFF", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of rivers",
       fill = "Sampling Year")
```

### Most common types and quantities of litter per region? (Group B: riverbanks; Group C: floating litter)

#### Group B: Riverside Waste

Statistics of the variable `RiversideWasteVariety_AttemptedSampling` to see how many samplings were performed.
```{r GroupB, warning=FALSE, message=FALSE}
Riverbank %>% 
  select(RiversideWasteVariety_AttemptedSampling) %>% 
  table()

```

To work with the `RiversideWasteVariety_` data, we need to filter the dataset to include only those rows where `RiversideWasteVariety_AttemptedSampling` is 1 (indicating that a sampling was attempted) and `RiversideWasteVariety_Qualified` is also 1 (indicating that the sampling was qualified).

```{r GroupBData, warning=FALSE, message=FALSE}
Riverbank %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1)
```
##### Percentage of qualified samplings

This represents the percentage of validated data among the total number of sampling beign attempted.
```{r}
Riverbank %>%
  summarise(
    total_rows = n(),
    total_valid = sum(
      RiversideWasteVariety_AttemptedSampling == 1 &
      RiversideWasteVariety_Qualified == 1,
      na.rm = TRUE
    ),
    pct_valid_total = round((total_valid / total_rows) * 100, digits = 1),
    pct_valid_percent = paste(round((total_valid / total_rows) * 100, 1), "%", sep = "")
  )
```

```{r}
validated_pct_by_country <- 
  Riverbank %>% 
  group_by(CountryResponsible) %>% 
  summarise(total_rows = n(),
            total_valid = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    `pct_valid_total(%)`        = round((total_valid / total_rows) * 100,1),
    total_attempts         = sum(RiversideWasteVariety_AttemptedSampling == 1, na.rm = TRUE),
    valid_among_attempts   = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    `pct_valid_among_attempts(%)` = round((valid_among_attempts / total_attempts) * 100,1), .groups = "drop") %>% 
  mutate(CountryResponsible = str_to_title(CountryResponsible))

validated_pct_by_country
```

```{r, fig.width= 10}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>% 
  summarise(
    n_attempted = sum(RiversideWasteVariety_AttemptedSampling == 1, na.rm = TRUE),
    n_validated = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    pct_validated = if_else(n_attempted > 0, n_validated / n_attempted * 100, 0),
    .groups = "drop"
  ) %>% 
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
  complete(CountryResponsible, SamplingYear,
           fill = list(n_attempted   = 0,
                       n_validated   = 0,
                       pct_validated = 0)
  ) %>% 
  filter(SamplingYear > 2021)  %>% 
  mutate(CountryResponsible = reorder(CountryResponsible,-pct_validated, FUN = mean)) %>% 
  ggplot(aes(
    x    = CountryResponsible,
    y    = pct_validated,
    fill = factor(SamplingYear)
  )) +
    geom_col(
      position = "dodge",
      colour  = "black"
    ) +
  geom_text(aes(label = n_attempted),
              position = position_dodge(width = 0.9),
              vjust    = -0.3,
              size     = 3
    ) +
  # geom_text(aes(label = paste0("(", n_attempted, ")")),
  #           position = position_dodge(width = 0.9),
  #           vjust    = -0.3,
  #           size     = 2.8) +
    scale_fill_manual(
      values = c(
        "2021" = "#E0FFFF",
        "2022" = "#B8D4F0",
        "2023" = "#4A90E2",
        "2024" = "#1E3A8A"
      ),
      name = "Sampling Year"
    ) +
    scale_y_continuous(
      labels = percent_format(scale = 1),
      expand = expansion(mult = c(0, 0.05))
    ) +
    labs(
      x     = NULL,
      y     = "Validated sampling (%)",
      title = "Percentage of validated records and total sampling per country and year"
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title  = element_text(hjust = 0.5)
    )
```

##### Most common types/quantities of litter

```{r, warning=FALSE, message=FALSE}

Riverbank %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1)

```
Firstly, we will pick up data to answer the question about group B. Only some variables are taken from the general dataset.

```{r, warning=FALSE, message=FALSE}
# GroupB dataset
GroupB_data <- 
  Riverbank %>% 
  select(DatasetID, CountryResponsible, SamplingCampaign, SamplingYear, SamplingCoordinatesLatitude, SamplingCoordinatesLongitude, SamplingRiverSystem, SamplingParticipants, RiversideWasteVariety_AttemptedSampling, RiversideWasteVariety_Qualified:RiversideWasteVariety_Results_ResultsComment)

```

The, Attributes' names are so complicated to work that I'll remote the preceding information **RiversideWasteVariety_Results_** which is common to all attributes; and finally to work with only the validated `Group B` dataset we create:

```{r, warning=FALSE, message=FALSE}
# GroupB dataset
GroupB_data_validated <- 
  Riverbank %>% 
  select(DatasetID, CountryResponsible, SamplingCampaign, SamplingYear, SamplingCoordinatesLatitude, SamplingCoordinatesLongitude, SamplingRiverSystem, SamplingParticipants, RiversideWasteVariety_AttemptedSampling, RiversideWasteVariety_Qualified:RiversideWasteVariety_Results_ResultsComment) %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1) %>% 
  # Primero quitamos `RiversideWasteVariety_Results_`
  rename_with(
    .cols = starts_with("RiversideWasteVariety_Results_"),
    # solo seleccionamos los tipos de residuos
    .fn = ~ gsub("RiversideWasteVariety_Results_", "", .) # Mantenemos la ultima extensión del nombre
  )

GroupB_data_validated
```
**CAUTION**!
There are some `NAs` values in the litter items that are probably incorrect since the groups did find other items. 
```{r}
summary(GroupB_data_validated)
```

We will obtain the plot for the most common (average abundance) litter type
```{r}
# Average litter type
Average_litter_type <- 
  GroupB_data_validated %>% 
  select(CountryResponsible, SamplingYear, PlasticBags:LocalWaste) %>%  
  pivot_longer(
    cols = PlasticBags:LocalWaste,
    names_to = "Litter_type",
    values_to = "Abundance") %>% 
  group_by(Litter_type) %>% 
  dplyr::summarise(Av_Abundance = mean(Abundance, na.rm = TRUE), 
                   SE_Abundance = sd(Abundance, na.rm = TRUE)/sqrt(n())) %>% 
  ggplot(., aes(x = reorder(Litter_type, -Av_Abundance), y = Av_Abundance)) +
  geom_errorbar(aes(ymin = Av_Abundance - SE_Abundance, ymax = Av_Abundance + SE_Abundance), width = 0.2) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0, 60), expand = c(0.01,0)) +
  #geom_text(aes(label = sprintf("%.1f", round(Av_Abundance, digits = 1))), vjust = 1.4, size = 3.5, colour = "white") +
  labs(x = "", y = "Litter abundance (mean)", title = '', subtitle = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

Average_litter_type
```


```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Average_Waste_Category_GroupB.png", plot = Average_litter_type, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 0.6, width = 35, height = 25, units = "cm", dpi = 150, limitsize = FALSE)
```


