---
title: "Plastic Pirates EU Analsysis"
author: "Dr. Antonio Canepa-Oneto & Dr. Patricia De La Fuente-Gamero"
date: "`r Sys.Date()`"
output:
  html_document: # rmarkdown::render("Plastic_Pirates_EU_Analysis.Rmd", "html_document")
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    keep_md: false
  md_document: # rmarkdown::render("Plastic_Pirates_EU_Analysis.Rmd", output_format = "md_document")
    variant: gfm 
---

# Data Analysis

## Loading Packages

Loading R packages necessary to perform the __EDA__.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(writexl)
library(DT)
library(lubridate)
library(scales)
library(vegan)       # decostand(), rda(), adonis2()
library(factoextra)  # fviz_pca_biplot(), fviz_dend()
library(randomForest)
library(RColorBrewer)
library(tidyHeatmap) # Requieres ComplexHeatmap --> BiocManager::install("ComplexHeatmap")
library(stringr)
```

## Data loading

```{r DataLoading, message=FALSE, warning=FALSE}
# Cargamos los datos incorporando las modificaciones para trabajar con él
# https://zenodo.org/records/15535434
# Cuidado con el nombre de los ríos, hay besòs y bessos.
Riverbank <- 
  read.csv(file = "PlasticPirates_Riverbank_data/Riverbank_data_v2.0.0.csv", header = TRUE, sep = ",", na.strings = "na") %>% 
  mutate(SamplingDate = lubridate::ymd(SamplingDate)) %>% 
  mutate(across(where(is.character), as.factor))
```

## Pre-proccessing

```{r RiversName, message=FALSE, warning=FALSE}
# Loading the rivers' dictionary
Riverbank_dictionary <- read_excel("Riverbank_data_v2.1.xlsx", sheet = "Final vocabulary clean")

Riverbank <-
  Riverbank %>%
  left_join(Riverbank_dictionary, by = "SamplingRiver") %>%
    #select(SamplingRiver, Corrected_river_English, SamplingRiverSystem, River_basin, everything()) %>% 
    mutate(River_basin = str_remove(River_basin, regex("\\s+Basin$", ignore_case = TRUE))) %>% 
  mutate(SamplingRiver = coalesce(Corrected_river_English, SamplingRiver),
         SamplingRiverSystem = River_basin) %>% 
  select(-c("No", "Country", "Corrected_river", "Corrected_river_English", "River_basin", "Count of SamplingRiver"))

```


## RESEARCH QUESTIONS

### Citizens Description

#### Numer of samplings per year and country

* Table 

```{r EDACountries, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n())
```

* Plot

```{r EDACountriesPlot, message=FALSE, warning=FALSE}
Temp_Num_Samp_Campaigns <- 
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Num_Sampling = n(), .groups = 'drop') %>%
    # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Num_Sampling = 0)) %>%
  # Ordenar por total acumulado
  mutate(CountryResponsible = reorder(CountryResponsible, -Num_Sampling, sum)) %>%
  ggplot(aes(x = CountryResponsible, y = Num_Sampling, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "dodge", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#E0FFFF", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of samplings campaigns",
       fill = "Sampling Year")

Temp_Num_Samp_Campaigns

#ggsave(filename = "Temporal_Number_Campaigns.png", plot = Temp_Num_Samp_Campaigns, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 1.3, width = 15, height = 10, units = "cm", dpi = 300, limitsize = FALSE)
```

#### Numer of Rivers sampled per year and country

*  Table

```{r EDARivers, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver))
```

* Plot (without numbers)

```{r EDARiversPlot, message=FALSE, warning=FALSE}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver), .groups = 'drop') %>%
  # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Rivers = 0)) %>%
  # Ordenar por total acumulado de ríos
  mutate(CountryResponsible = reorder(CountryResponsible, -Rivers, sum)) %>%
  ggplot(aes(x = CountryResponsible, y = Rivers, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = "dodge", colour = 'black') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # Paleta de colores con intensidad creciente hacia 2024
  scale_fill_manual(values = c("2021" = "#E0FFFF", "2022" = "#B8D4F0", "2023" = "#4A90E2", "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "", 
       y = "Number of rivers",
       fill = "Sampling Year")
```

* Plot (with numbers)

```{r EDARiversPlotNum, message=FALSE, warning=FALSE}
# Preparar los datos
plot_data <- 
  Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>%
  summarise(Rivers = n_distinct(SamplingRiver), .groups = 'drop') %>%
  # Capitalizar primera letra de cada país
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>%
  # Crear todas las combinaciones posibles de país y año
  complete(CountryResponsible, SamplingYear, fill = list(Rivers = 0)) %>%
  # Ordenar por total acumulado de ríos
  mutate(CountryResponsible = reorder(CountryResponsible, -Rivers, sum))

# Crear el gráfico
Num_rios_year <- 
ggplot(plot_data, aes(x = CountryResponsible, y = Rivers, fill = factor(SamplingYear))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), colour = "black") +
# Etiquetas encima de cada barra
  geom_text(aes(label = Rivers), position = position_dodge(width = 0.9), vjust = -0.3, # coloca el texto encima
    color = "black", # color negro como quieres 
    fontface = "bold", size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  scale_fill_manual(values = c(
    "2021" = "#E0FFFF",
    "2022" = "#B8D4F0",
    "2023" = "#4A90E2",
    "2024" = "#1E3A8A"
  )) +
# Más espacio arriba para que no se corten las etiquetas
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "", y = "Number of rivers", fill = "Sampling Year")

print(Num_rios_year)

#ggsave(filename = "Temporal_Number_Rivers.png", plot = Num_rios_year, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 1.9, width = 15, height = 10, units = "cm", dpi = 300, limitsize = FALSE)
```

### Most common types and quantities of litter per region? (Group B: riverbanks; Group C: floating litter)

#### Group B: Riverside Waste

Statistics of the variable `RiversideWasteVariety_AttemptedSampling` to see how many samplings were performed.
```{r GroupB, warning=FALSE, message=FALSE, eval=TRUE}
Riverbank %>% 
  select(RiversideWasteVariety_AttemptedSampling) %>% 
  table()
```

To work with the `RiversideWasteVariety_` data, we need to filter the dataset to include only those rows where `RiversideWasteVariety_AttemptedSampling` is 1 (indicating that a sampling was attempted) and `RiversideWasteVariety_Qualified` is also 1 (indicating that the sampling was qualified).

```{r GroupBData, warning=FALSE, message=FALSE, eval=FALSE}
Riverbank %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1)
```

##### Percentage of qualified samplings

This represents the percentage of validated data among the total number of sampling beign attempted.
```{r}
Riverbank %>%
  summarise(
    total_rows = n(),
    total_valid = sum(
      RiversideWasteVariety_AttemptedSampling == 1 &
      RiversideWasteVariety_Qualified == 1,
      na.rm = TRUE
    ),
    pct_valid_total = round((total_valid / total_rows) * 100, digits = 1),
    pct_valid_percent = paste(round((total_valid / total_rows) * 100, 1), "%", sep = "")
  )
```

```{r}
validated_pct_by_country <- 
  Riverbank %>% 
  group_by(CountryResponsible) %>% 
  summarise(total_rows = n(),
            total_valid = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    `pct_valid_total(%)`        = round((total_valid / total_rows) * 100,1),
    total_attempts         = sum(RiversideWasteVariety_AttemptedSampling == 1, na.rm = TRUE),
    valid_among_attempts   = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    `pct_valid_among_attempts(%)` = round((valid_among_attempts / total_attempts) * 100,1), .groups = "drop") %>% 
  mutate(CountryResponsible = str_to_title(CountryResponsible))

validated_pct_by_country
```

```{r, fig.width= 10}
Riverbank %>% 
  group_by(CountryResponsible, SamplingYear) %>% 
  summarise(
    n_attempted = sum(RiversideWasteVariety_AttemptedSampling == 1, na.rm = TRUE),
    n_validated = sum(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1, na.rm = TRUE),
    pct_validated = if_else(n_attempted > 0, n_validated / n_attempted * 100, 0),
    .groups = "drop"
  ) %>% 
  mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
  complete(CountryResponsible, SamplingYear,
           fill = list(n_attempted   = 0,
                       n_validated   = 0,
                       pct_validated = 0)
  ) %>% 
  filter(SamplingYear > 2021)  %>% 
  mutate(CountryResponsible = reorder(CountryResponsible,-pct_validated, FUN = mean)) %>% 
  ggplot(aes(
    x    = CountryResponsible,
    y    = pct_validated,
    fill = factor(SamplingYear)
  )) +
    geom_col(
      position = "dodge",
      colour  = "black"
    ) +
  geom_text(aes(label = paste0("(",n_attempted,")")),
              position = position_dodge(width = 0.9),
              vjust    = -0.3,
              size     = 3
    ) +
  # geom_text(aes(label = paste0("(", n_attempted, ")")),
  #           position = position_dodge(width = 0.9),
  #           vjust    = -0.3,
  #           size     = 2.8) +
    scale_fill_manual(
      values = c(
        "2021" = "#E0FFFF",
        "2022" = "#B8D4F0",
        "2023" = "#4A90E2",
        "2024" = "#1E3A8A"
      ),
      name = "Sampling Year"
    ) +
    scale_y_continuous(
      labels = percent_format(scale = 1),
      expand = expansion(mult = c(0, 0.05))
    ) +
    labs(
      x     = NULL,
      y     = "Validated sampling (%)",
      title = "Percentage of validated records (sampling events) per country and year"
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title  = element_text(hjust = 0.5)
    )
```

##### Most common types/quantities of litter

```{r, warning=FALSE, message=FALSE, eval=FALSE}

Riverbank %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1)

```
Firstly, we will pick up data to answer the question about group B. Only some variables are taken from the general dataset.

```{r, warning=FALSE, message=FALSE}
# GroupB dataset
GroupB_data <- 
  Riverbank %>% 
  select(DatasetID, CountryResponsible, SamplingCampaign, SamplingYear, SamplingCoordinatesLatitude, SamplingCoordinatesLongitude, SamplingRiverSystem, SamplingParticipants, RiversideWasteVariety_AttemptedSampling, RiversideWasteVariety_Qualified:RiversideWasteVariety_Results_ResultsComment)

```

The, Attributes' names are so complicated to work that I'll remote the preceding information **RiversideWasteVariety_Results_** which is common to all attributes; and finally to work with only the validated `Group B` dataset we create:

```{r, warning=FALSE, message=FALSE}
# GroupB dataset
GroupB_data_validated <- 
  Riverbank %>% 
  select(DatasetID, CountryResponsible, SamplingCampaign, SamplingYear, SamplingCoordinatesLatitude, SamplingCoordinatesLongitude, SamplingRiverSystem, SamplingParticipants, RiversideWasteVariety_AttemptedSampling, RiversideWasteVariety_Qualified:RiversideWasteVariety_Results_ResultsComment) %>% 
  filter(RiversideWasteVariety_AttemptedSampling == 1 & RiversideWasteVariety_Qualified == 1) %>% 
  # Primero quitamos `RiversideWasteVariety_Results_`
  rename_with(
    .cols = starts_with("RiversideWasteVariety_Results_"),
    # solo seleccionamos los tipos de residuos
    .fn = ~ gsub("RiversideWasteVariety_Results_", "", .) # Mantenemos la ultima extensión del nombre
  )

#GroupB_data_validated

# Después de la reunión del día 20260212, decidimos que hay que quitar la variable `SingleUsePlasticTotal`.
GroupB_data_validated <-
  GroupB_data_validated %>% 
  select(!SingleUsePlasticTotal)
```

**CAUTION**!
There are some `NAs` values in the litter items that are probably incorrect since the groups did find other items. 
```{r, eval=FALSE}
summary(GroupB_data_validated)
```

After meeting and email from Charlotte, we realise that NA in group B for some items it's really a zero. So we update these columns replacing `NA` values with a `0`.

```{r}
GroupB_data_validated <-
  GroupB_data_validated %>% 
  mutate(across(PlasticBags:LocalWaste, ~replace_na(., 0)))
```


###### Litter type abundance

We will obtain the plot for the most common (average abundance) litter type
```{r}
# Average litter type
Average_litter_type <- 
  GroupB_data_validated %>% 
  select(CountryResponsible, SamplingYear, PlasticBags:LocalWaste) %>%  
  pivot_longer(
    cols = PlasticBags:LocalWaste,
    names_to = "Litter_type",
    values_to = "Abundance") %>% 
  group_by(Litter_type) %>% 
  dplyr::summarise(Av_Abundance = mean(Abundance, na.rm = TRUE), 
                   SE_Abundance = sd(Abundance, na.rm = TRUE)/sqrt(n())) %>% 
  ggplot(., aes(x = reorder(Litter_type, -Av_Abundance), y = Av_Abundance)) +
  geom_errorbar(aes(ymin = Av_Abundance - SE_Abundance, ymax = Av_Abundance + SE_Abundance), width = 0.2) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0, 30), expand = c(0.01,0)) +
  #geom_text(aes(label = sprintf("%.1f", round(Av_Abundance, digits = 1))), vjust = 1.4, size = 3.5, colour = "white") +
  labs(x = "",  y = expression(Litter~abundance~"(mean" %+-% s.e.~")"), title = '', subtitle = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

print(Average_litter_type)
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Average_Waste_Category_GroupB.png", plot = Average_litter_type, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 1.3, width = 15, height = 10, units = "cm", dpi = 300, limitsize = FALSE)
```


###### Litter type abundance per country

If we now want to visualize the contribution per each country, we have:

```{r AverageLitterCountry, message=FALSE, warning=FALSE}
# Average litter type
Average_litter_type_country <- 
GroupB_data_validated %>% 
  select(CountryResponsible, PlasticBags:LocalWaste) %>%  
  pivot_longer(
    PlasticBags:LocalWaste,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>% 
  # 1) group by country AND type
  group_by(CountryResponsible, Litter_type) %>% 
  summarise(
    Av_Abundance = mean(Abundance, na.rm = TRUE),
    SE_Abundance = sd(Abundance,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>% 
  # title‐case country, reorder types by their overall mean if you like
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    Litter_type       = reorder(Litter_type, -Av_Abundance)
  ) %>% 
  # 2) plot bars first, then error bars
  ggplot(aes(x = Litter_type, y = Av_Abundance)) +
  #geom_col(fill = "lightgray", colour = "black") +
  geom_errorbar(aes(ymin = Av_Abundance - SE_Abundance,
                    ymax = Av_Abundance + SE_Abundance), width = 0.2) +
  geom_col(fill = "lightgray", colour = "black") +
  facet_wrap(~ CountryResponsible, ncol   = 3, scales = "free_y") +
  labs(x = "", y = expression(Litter~abundance~"(mean" %+-% s.e.~")"), title = '', subtitle = "") +
  # theme_classic() +
  # theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  theme_classic(base_size = 16) +
  theme(
    # rotate axis labels, center title, etc.
    axis.text.x     = element_text(angle = 65, hjust = 1),
    plot.title      = element_text(hjust = 0.5),
    strip.background = element_rect(
      fill   = "lightgrey",  # background color
      color  = "black",      # border color
      size   = 0.5           # border thickness
    ),
    strip.text       = element_text(
      colour = "black",      # text color
      face   = "bold"        # make it stand out
    )
  )

Average_litter_type_country
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Average_Waste_Category_Country_GroupB.png", plot = Average_litter_type_country, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 1, width = 40, height = 30, units = "cm", dpi = 300, limitsize = FALSE)
```

###### Litter type Proportion I - "Across‑type" proportions

* The first approach will be a "*Across‑type*" proportions. 
  * For each litter type, what fraction comes from each country? (not from an european total)
  * total abundance of that litter type over each total country.


```{r ProportionLitterCountry, message=FALSE, warning=FALSE}
Prop_Liter_Country <- 
  GroupB_data_validated %>%
  # 1) pivot to long form
  pivot_longer(
    cols      = PlasticBags:LocalWaste,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  # 2) sum abundance by (Country, Type)
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 3) compute, per type, the global total
  group_by(Litter_type) %>%
  mutate(
    Global_Total   = sum(Total_Abundance),
    Prop_Country   = Total_Abundance / Global_Total
  ) %>%
  ungroup()

Prop_Liter_Country
```


```{r ProportionLitterCountryPlot, message=FALSE, warning=FALSE}
# Example 100%-stacked bar
Country_contribution_litter_type <- 
Prop_Liter_Country %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = Litter_type, y = Prop_Country, fill = CountryResponsible)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Country"
  ) +
  labs(
    x = "Litter type",
    y = "Relative abundance (%)",
    title = "Country contributions to each litter type"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  coord_flip()

print(Country_contribution_litter_type)
```
```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Country_contribution_litter_type.png", plot = Country_contribution_litter_type, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 0.5, width = 40, height = 30, units = "cm", dpi = 300, limitsize = FALSE)
```


* Now, considering the total of Europe

```{r ProportionLitterEurope, warning=FALSE, message=FALSE}
# 1) Pivot, sum by country & type, then compute each country’s share of the European total
Data_prop_eu <- 
  GroupB_data_validated %>%
  pivot_longer(
    cols      = PlasticBags:LocalWaste,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Litter_type) %>%
  mutate(
    European_Total   = sum(Total_Abundance),
    Prop_of_Europe   = Total_Abundance / European_Total
  ) %>%
  ungroup() %>%
  # optional: title‑case and reorder types by overall Prop for consistent ordering
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    Litter_type        = reorder(Litter_type, -Prop_of_Europe)
  )

Data_prop_eu
```

To compare and validate the output, for each `litter_type` the `European_Total` have to be the same.

```{r ProportionLitterEuropeValidation, warning=FALSE, message=FALSE}
# To compare and validate the output, for each `litter_type` the `European_Total` have to be the same.
Data_prop_eu %>% 
  filter(Litter_type == "AluminiumFoil")
```

```{r ProportionLitterEuropePlotStacked, warning=FALSE, message=FALSE, out.width='100%', fig.align='center'}
# Example 100%-stacked bar
Data_prop_eu %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = Litter_type, y = Prop_of_Europe, fill = CountryResponsible)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Country"
  ) +
  labs(
    x = "Litter type",
    y = "Share of global abundance",
    title = "Country contributions to each litter type (Europe as basis)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  coord_flip()
```


```{r ProportionLitterEuropePlot, warning=FALSE, message=FALSE, out.width='50%', fig.align='center'}
#x11()
# 2) Faceted bar chart of proportion (country share of European total)
# 2) plot with fixed scales and text labels
ggplot(Data_prop_eu, aes(x = Litter_type, y = Prop_of_Europe)) +
  geom_col(fill = "lightgray", colour = "black") +
  geom_text( # comment if you don't need the numbers above the bars
    #aes(label = percent(Prop_of_Europe, accuracy = 0.1), suffix = "%"),
    aes(label = round((Prop_of_Europe*100), 1)),
    angle   = 45,       # rotate labels 45°
    hjust   = 0,        # left‑align text relative to its anchor
    vjust   = -0.3,     # nudge it a bit above the bar
    size  = 2.7          # adjust text size as needed
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.15))  # extra 15% headroom for labels
  ) +
  facet_wrap(~ CountryResponsible, ncol = 3, scales = "fixed") +
  labs(
    x        = NULL,
    y        = "Share of European total (%)",
    title    = "Country contributions to each litter type",
    subtitle = "Proportion of European total abundance for each category"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x      = element_text(angle = 65, hjust = 1),
    strip.background = element_rect(fill = "lightgrey", colour = "black"),
    strip.text       = element_text(face = "bold")
  )
```

* HeatMaps

```{r ProportionLitterEuropeHeatMap, warning=FALSE, message=FALSE, out.width='50%', fig.align='center'}
ProportionLitterEuropeHeatMap <- 
Data_prop_eu %>%
  heatmap(
    .row           = Litter_type,
    .column        = CountryResponsible,
    .value         = Prop_of_Europe,
    scale          = "column", # can be "none" (by default) or "row", "column" or "both". (#column keeps the z‑scaling)
    # a three‑colour gradient: low→mid→high
    palette_value  = c("#F7FBFF", "#6BAED6", "#08306B"),
    name           = "Relative \nimportance (%)",
    column_title   = "Country",
    row_title      = "Litter type",
    show_row_names = TRUE,
    show_column_names = TRUE
  )

print(ProportionLitterEuropeHeatMap)
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Proportion_Litter_Europe_HeatMap.png", width = 25, height = 20, units = "cm", res = 300, pointsize = 1)
ProportionLitterEuropeHeatMap
dev.off()
```

###### Litter type Proportion II - "Within‑country" proportions

* The second approach will be a "*Within‑country*" proportions. 
  * For each country, what fraction of its total litter is each type?
  * Total abundance of all types in that country.


```{r ProportionLitterCountryII, message=FALSE, warning=FALSE}
Prop_Liter_Country2 <- 
  GroupB_data_validated %>%
  # 1) pivot to long
  pivot_longer(
    cols      = PlasticBags:LocalWaste,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  # 2) sum by (Country, Type)
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 3) compute, per country, that country’s total
  group_by(CountryResponsible) %>%
  mutate(
    Country_Total = sum(Total_Abundance),
    Prop_Type     = Total_Abundance / Country_Total
  ) %>%
  ungroup()

Prop_Liter_Country2
```


```{r ProportionLitterCountryIIPlot, message=FALSE, warning=FALSE}
# Example 100%-stacked bar per country
Prop_Liter_Country2 %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = CountryResponsible, y = Prop_Type, fill = Litter_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Country",
    y = "Share of country’s total",
    title = "Composition of litter types within each country"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))
```


```{r ProportionLitterCountryIIPlot2, message=FALSE, warning=FALSE}
  # 2) plot bars first, then error bars
Prop_Liter_Country2 %>% 
  # 1) group by country AND type
  group_by(CountryResponsible, Litter_type) %>% 
  summarise(
    Av_Proportion = mean(Prop_Type, na.rm = TRUE),
    SE_Proportion = sd(Prop_Type,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>% 
  # title‐case country, reorder types by their overall mean if you like
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    Litter_type       = reorder(Litter_type, -Av_Proportion)
  ) %>% 
  ggplot(data = ., aes(x = Litter_type, y = Av_Proportion)) +
  geom_col(fill = "lightgray", colour = "black") +
  geom_errorbar(aes(ymin = Av_Proportion - SE_Proportion,
                    ymax = Av_Proportion + SE_Proportion), width = 0.2) +
  facet_wrap(~ CountryResponsible, ncol   = 3, scales = "free_y") +
  labs(x = "", y = "Litter abundance (mean)", title = '', subtitle = "") +
  # theme_classic() +
  # theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  theme_classic(base_size = 16) +
  theme(
    # rotate axis labels, center title, etc.
    axis.text.x     = element_text(angle = 65, hjust = 1),
    plot.title      = element_text(hjust = 0.5),
    strip.background = element_rect(
      fill   = "lightgrey",  # background color
      color  = "black",      # border color
      size   = 0.5           # border thickness
    ),
    strip.text       = element_text(
      colour = "black",      # text color
      face   = "bold"        # make it stand out
    )
  )

```

###### Multivariate Perspective

```{r DataPreparationMultivariate}
# assume your data is in a data.frame called `df`
# and has at least:
#   CountryResponsible | PlasticBags | PlasticBottles | ... | LocalWaste
# Replace the NA values for 0. --> Probably the default behaviour?
df <- GroupB_data_validated %>% 
  select(CountryResponsible, PlasticBags:LocalWaste)
# %>% 
# #drop_na()
# mutate_all(funs(replace_na(.,0)))

# isolate metadata + community matrix
meta <- df %>% select(CountryResponsible)
comm <- df %>% select(-CountryResponsible)

# Hellinger‐transform community data (recommended for PCA/cluster/PERMANOVA)
comm_hel <- decostand(comm, method = "hellinger")

```


In order to see if there are significant differences across country in the community composition, we will perform a PERMANOVA test:

```{r PermanovaTest, message=FALSE, warning=FALSE}
# 2a) Compute distance and hierarchical clustering (Ward’s D2)
dist_mat <- vegdist(comm_hel, method = "euclid")

# 3a) PERMANOVA: test if composition differs by Country
set.seed(42)

perm1 <- adonis2(dist_mat ~ CountryResponsible, data = meta, 
                 permutations = 999, by = "margin")
print(perm1)
```

So, we conclude that there are significant differences among countries. but What are the items that explains these differences? 

The answer will be addressed by performing a **Multivariate Random Forest**.

```{r MultRandFor, message=FALSE, warning=FALSE}
# 3b) Multivariate Random Forest:  
#     predict Country from composition → variable importance
rf <- randomForest(
  x      = comm_hel,
  y      = as.factor(meta$CountryResponsible),
  ntree  = 500,
  importance = TRUE
)

print(rf)
```
The estimated error rate (OOB) is high, so conclusion have to be taken with care. Thus, the more important (differential across countries) litter components are:

```{r RFVarImport, message=FALSE, warning=FALSE}
varImpPlot(rf, main="Random Forest: Litter‐type Importance for Country Discrimination")
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Random_Forest_Importance.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
varImpPlot(rf, main="Random Forest: Litter‐type Importance for Country Discrimination")
dev.off()
```

This plot of variable importance importance is telling us:

*  A large `MeanDecreaseAccuracy` (**MDA**) means that, when you break the relationship between that predictor and the response, the model’s accuracy suffers a lot → it’s an important predictor for correct country assignments. 
  * **MDA** is generally the more reliable gauge of predictive importance, since it directly measures impact on OOB‐accuracy.

* A large `MeanDecreaseGini` (**MDG**) means that X frequently appears in high‑up the tree splits and yields big purity gains → it’s structurally important to partitioning the data. 
  * **MDG** is faster to compute and tells you which features the trees “prefer” splitting on, but can overweight high–cardinality or continuous variables.
  
Finally, we will perform a Principal Component Analysis (PCA) to see correlation among litter items and *possible* associations with countries

```{r PCA, warning=FALSE, message=FALSE}
# ─── 1) PCA (and optional RDA constrained by Country) ───────────────────────────
# 1a) Unconstrained PCA on Hellinger data
# pca1 <- rda(comm_hel)  
# 
# summary(pca1)
# 
# plot(pca1)

# (re)compute PCA with prcomp
pca_prcomp <- prcomp(comm_hel, scale. = TRUE)

#summary(pca_prcomp)
```


```{r PCAplot, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# 1b) Quick biplot via factoextra
PCA_Litter_Country <- 
fviz_pca_biplot(pca_prcomp,
                geom.ind    = "point",            # show sites as points
                col.ind     = meta$CountryResponsible %>% str_to_title(), 
                palette.ind = "jco",              # color‐blind‐friendly palette
                addEllipses = FALSE,               # 95% CI per country
                label      = "var",               # show arrows for variables
                repel      = TRUE
) +
  labs(title = "PCA of Litter Composition (Hellinger‐transformed)",
       subtitle = "Points = samples colored by Country; arrows = litter types") +
  theme_minimal(base_size = 14)

PCA_Litter_Country
```


```{r PCAplot2, warning=FALSE, message=FALSE}
# Para el gráfico calcularemos los centroides y los pintaremos de manera diferencial

# Calcular centroides manualmente
scores <- as.data.frame(pca_prcomp$x)
scores$Country <- meta$CountryResponsible %>% str_to_title()

centroids <- 
  scores %>%
  group_by(Country) %>%
  summarise(PC1 = mean(PC1),
            PC2 = mean(PC2))

# Definir niveles y paleta de colores para puntos y centroides, una sola vez
meta$Country <- factor(str_to_title(meta$CountryResponsible))

# Definir niveles únicos y ordenados
countries <- levels(meta$Country)

# Crear paleta manual coherente
my_palette <- setNames(
  brewer.pal(length(countries), "Set3"),
  countries
)

# PCA con puntos pequeños (color)
p <- fviz_pca_biplot(
  pca_prcomp,
  geom.ind   = "point",
  col.ind    = meta$Country,
  addEllipses = FALSE,
  label      = "var",
  repel      = TRUE,
  pointsize  = 2.5,
  pointshape = 16,
  alpha.ind  = 0.6
)

# Añadir centroides (fill)
# scores <- as.data.frame(pca_prcomp$x) # Esto ya estaba
# scores$Country <- meta$Country # Esto ya estaba
# 
# centroids <- 
#   scores %>% 
#   dplyr::group_by(Country) %>% 
#   dplyr::summarise(PC1 = mean(PC1),
#                    PC2 = mean(PC2))


PCA_Litter_Country2 <- 
# x11()
p +
  geom_point(data = centroids,
             aes(x = PC1, y = PC2, fill = Country),
             shape = 21,
             size = 5,
             color = "black",
             stroke = 1.2) +
  
  scale_color_manual(values = my_palette, drop = FALSE) +
  scale_fill_manual(values = my_palette, drop = FALSE) +
  
  labs(title = "PCA of Litter Composition (Hellinger‐transformed)",
       subtitle = "Small points = samples; Large points = country centroids",
       color = "Country",
       fill = "Country") +
  
  theme_minimal(base_size = 14)

PCA_Litter_Country2
```



```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/PCA_Litter_Country.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
PCA_Litter_Country2
dev.off()
```

And the most important Items for the two forst principal components are:

* First PC
```{r PC1VarImport, message=FALSE, warning=FALSE}
# Bar‐plot of top 10 contributors to PC1
fviz_contrib(pca_prcomp, choice = "var", axes = 1, top = 10) +
  ggtitle("Top 10 variables contributing to PC1")
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Top10_PC1_PCA_Litter_Country.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
# Bar‐plot of top 10 contributors to PC1
fviz_contrib(pca_prcomp, choice = "var", axes = 1, top = 10) +
  ggtitle("Top 10 variables contributing to PC1")
dev.off()
```


```{r PC2VarImport, message=FALSE, warning=FALSE}
# Bar‐plot of top 10 contributors to PC2
fviz_contrib(pca_prcomp, choice = "var", axes = 2, top = 10) +
  ggtitle("Top 10 variables contributing to PC2")
```
```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Top10_PC2_PCA_Litter_Country.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
# Bar‐plot of top 10 contributors to PC2
fviz_contrib(pca_prcomp, choice = "var", axes = 2, top = 10) +
  ggtitle("Top 10 variables contributing to PC2")
dev.off()
```

#### Group B: Single Use Plastic - Riverside Waste

Cogeremos el subset de datos que si representan los plásticos de un solo uso (`SUP`) para el grupo B.

```{r, warning=FALSE, message=FALSE}
# GroupB dataset
GroupB_SingleUsePlastic_data_validated <- 
  GroupB_data_validated %>% 
  select(DatasetID:Polystyrene) # Seleccionamos metadata y los Single Use Plastic

GroupB_SingleUsePlastic_data_validated

```

```{r}
summary(GroupB_SingleUsePlastic_data_validated)
```


###### SUP type abundance

We will obtain the plot for the most common (average abundance) Single Use Plastic type
```{r}
# Average litter type
SUP_Average_litter_type <- 
  GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, SamplingYear, PlasticBags:Polystyrene) %>%  
  pivot_longer(
    cols = PlasticBags:Polystyrene,
    names_to = "Litter_type",
    values_to = "Abundance") %>% 
  group_by(Litter_type) %>% 
  dplyr::summarise(Av_Abundance = mean(Abundance, na.rm = TRUE), 
                   SE_Abundance = sd(Abundance, na.rm = TRUE)/sqrt(n())) %>% 
  ggplot(., aes(x = reorder(Litter_type, -Av_Abundance), y = Av_Abundance)) +
  geom_errorbar(aes(ymin = Av_Abundance - SE_Abundance, ymax = Av_Abundance + SE_Abundance), width = 0.2) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0, 20), expand = c(0.01,0)) +
  #geom_text(aes(label = sprintf("%.1f", round(Av_Abundance, digits = 1))), vjust = 1.4, size = 3.5, colour = "white") +
  labs(x = "", y = expression("Single use plastic abundance (mean" %+-% "s.e.)"), title = '', subtitle = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

print(SUP_Average_litter_type)
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Average_SUP_GroupB.png", plot = SUP_Average_litter_type, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 0.6, width = 35, height = 25, units = "cm", dpi = 150, limitsize = FALSE)
```

###### SUP type abundance per country

If we now want to visualize the contribution per each country, we have:

```{r AverageLitterCountrySUP, message=FALSE, warning=FALSE}
# Average litter type
SUP_Average_litter_type_country <- 
GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, PlasticBags:Polystyrene) %>%  
  pivot_longer(
    PlasticBags:Polystyrene,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>% 
  # 1) group by country AND type
  group_by(CountryResponsible, Litter_type) %>% 
  summarise(
    Av_Abundance = mean(Abundance, na.rm = TRUE),
    SE_Abundance = sd(Abundance,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>% 
  # title‐case country, reorder types by their overall mean if you like
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    #Litter_type       = reorder(Litter_type, -Av_Abundance) # Reordenando el eje X (poco util)
    Litter_type       = Litter_type
  ) %>% 
  # 2) plot bars first, then error bars
  ggplot(aes(x = reorder(Litter_type, -Av_Abundance), y = Av_Abundance)) +
  geom_errorbar(aes(ymin = Av_Abundance - SE_Abundance,
                    ymax = Av_Abundance + SE_Abundance), width = 0.2) +
  geom_col(fill = "lightgray", colour = "black") +
  facet_wrap(~ CountryResponsible, ncol   = 3, scales = "free_y") +
  labs(x = "", y = expression("Single use plastic abundance (mean" %+-% "s.e.)"), title = '', subtitle = "") +
  # theme_classic() +
  # theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  theme_classic(base_size = 16) +
  theme(
    # rotate axis labels, center title, etc.
    axis.text.x     = element_text(angle = 65, hjust = 1),
    plot.title      = element_text(hjust = 0.5),
    strip.background = element_rect(
      fill   = "lightgrey",  # background color
      color  = "black",      # border color
      size   = 0.5           # border thickness
    ),
    strip.text       = element_text(
      colour = "black",      # text color
      face   = "bold"        # make it stand out
    )
  )

SUP_Average_litter_type_country
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Average_SUP_Country_GroupB.png", plot = SUP_Average_litter_type_country, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 1, width = 40, height = 30, units = "cm", dpi = 150, limitsize = FALSE)
```


###### SUP type Proportion I - "Across‑type" proportions

* The first approach will be a "*Across‑type*" proportions. 
  * For each litter type, what fraction comes from each country? (not from an european total)
  * total abundance of that litter type over each total country.


```{r ProportionLitterCountrySUP, message=FALSE, warning=FALSE}
SUP_Prop_Liter_Country <- 
  GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, PlasticBags:Polystyrene) %>%  
  pivot_longer(
    PlasticBags:Polystyrene,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  # 2) sum abundance by (Country, Type)
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 3) compute, per type, the global total
  group_by(Litter_type) %>%
  mutate(
    Global_Total   = sum(Total_Abundance),
    Prop_Country   = Total_Abundance / Global_Total
  ) %>%
  ungroup()

SUP_Prop_Liter_Country
```


```{r ProportionLitterCountryPlotSUP, message=FALSE, warning=FALSE}
# Example 100%-stacked bar
Country_contribution_SUP_type <- 
SUP_Prop_Liter_Country %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = Litter_type, y = Prop_Country, fill = CountryResponsible)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Country"
  ) +
  labs(
    x = "Single use plastic type",
    y = "Relative abundance (%)",
    title = "Country contributions to each litter type"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  coord_flip()

print(Country_contribution_SUP_type)
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
ggsave(filename = "Country_contribution_SUP_type.png", plot = Country_contribution_SUP_type, path = paste(getwd(), "/OUTPUT/FIGURES", sep = ""), scale = 0.5, width = 40, height = 30, units = "cm", dpi = 300, limitsize = FALSE)
```

* Now, considering the total of Europe

```{r ProportionLitterEuropeSUP, warning=FALSE, message=FALSE}
# 1) Pivot, sum by country & type, then compute each country’s share of the European total
SUP_Data_prop_eu <- 
  GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, PlasticBags:Polystyrene) %>%  
  pivot_longer(
    PlasticBags:Polystyrene,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Litter_type) %>%
  mutate(
    European_Total   = sum(Total_Abundance),
    Prop_of_Europe   = Total_Abundance / European_Total
  ) %>%
  ungroup() %>%
  # optional: title‑case and reorder types by overall Prop for consistent ordering
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    Litter_type        = reorder(Litter_type, -Prop_of_Europe)
  )

SUP_Data_prop_eu
```

```{r ProportionLitterEuropePlotStackedSUP, warning=FALSE, message=FALSE, out.width='100%', fig.align='center'}
# Example 100%-stacked bar
SUP_Data_prop_eu %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = Litter_type, y = Prop_of_Europe, fill = CountryResponsible)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Country"
  ) +
  labs(
    x = "Single use plastic type",
    y = "Share of global abundance",
    title = "Country contributions to each litter type (Europe as basis)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  coord_flip()
```


```{r ProportionLitterEuropePlotSUP, warning=FALSE, message=FALSE, out.width='50%', fig.align='center'}
#x11()
# 2) Faceted bar chart of proportion (country share of European total)
# 2) plot with fixed scales and text labels
ggplot(SUP_Data_prop_eu, aes(x = Litter_type, y = Prop_of_Europe)) +
  geom_col(fill = "lightgray", colour = "black") +
  geom_text( # comment if you don't need the numbers above the bars
    #aes(label = percent(Prop_of_Europe, accuracy = 0.1), suffix = "%"),
    aes(label = round((Prop_of_Europe*100), 1)),
    angle   = 45,       # rotate labels 45°
    hjust   = 0,        # left‑align text relative to its anchor
    vjust   = -0.3,     # nudge it a bit above the bar
    size  = 2.7          # adjust text size as needed
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.15))  # extra 15% headroom for labels
  ) +
  facet_wrap(~ CountryResponsible, ncol = 3, scales = "fixed") +
  labs(
    x        = NULL,
    y        = "Share of European total (%)",
    title    = "Country contributions to each litter type",
    subtitle = "Proportion of European total abundance for each category"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x      = element_text(angle = 65, hjust = 1),
    strip.background = element_rect(fill = "lightgrey", colour = "black"),
    strip.text       = element_text(face = "bold")
  )
```

* HeatMaps

```{r ProportionLitterEuropeHeatMapSUP, warning=FALSE, message=FALSE, out.width='50%', fig.align='center'}
ProportionSUPEuropeHeatMap <- 
SUP_Data_prop_eu %>%
  heatmap(
    .row           = Litter_type,
    .column        = CountryResponsible,
    .value         = Prop_of_Europe,
    scale          = "column", # can be "none" (by default) or "row", "column" or "both". (#column keeps the z‑scaling)
    # a three‑colour gradient: low→mid→high
    palette_value  = c("#F7FBFF", "#6BAED6", "#08306B"),
    name           = "Relative \nimportance (%)",
    column_title   = "Country",
    row_title      = "Litter type",
    show_row_names = TRUE,
    show_column_names = TRUE
  )

print(ProportionSUPEuropeHeatMap)
```
```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Proportion_SUP_Europe_HeatMap.png", width = 25, height = 20, units = "cm", res = 300, pointsize = 1)
ProportionSUPEuropeHeatMap
dev.off()
```

###### SUP type Proportion II - "Within‑country" proportions

* The second approach will be a "*Within‑country*" proportions. 
  * For each country, what fraction of its total litter is each type?
  * Total abundance of all types in that country.


```{r ProportionLitterCountryIISUP, message=FALSE, warning=FALSE}
SUP_Prop_Liter_Country2 <- 
  GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, PlasticBags:Polystyrene) %>%  
  pivot_longer(
    PlasticBags:Polystyrene,
    names_to  = "Litter_type",
    values_to = "Abundance"
  ) %>%
  # 2) sum by (Country, Type)
  group_by(CountryResponsible, Litter_type) %>%
  summarise(
    Total_Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 3) compute, per country, that country’s total
  group_by(CountryResponsible) %>%
  mutate(
    Country_Total = sum(Total_Abundance),
    Prop_Type     = Total_Abundance / Country_Total
  ) %>%
  ungroup()

SUP_Prop_Liter_Country2
```


```{r ProportionLitterCountryIIPlotSUP, message=FALSE, warning=FALSE}
# Example 100%-stacked bar per country
SUP_Prop_Liter_Country2 %>% 
mutate(CountryResponsible = str_to_title(CountryResponsible)) %>% 
ggplot(., aes(x = CountryResponsible, y = Prop_Type, fill = Litter_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Country",
    y = "Share of country’s total",
    title = "Composition of litter types within each country",
    fill = "Single use \nplastic type"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))
```


```{r ProportionLitterCountryIIPlot2SUP, message=FALSE, warning=FALSE}
  # 2) plot bars first, then error bars
SUP_Prop_Liter_Country2 %>% 
  # 1) group by country AND type
  group_by(CountryResponsible, Litter_type) %>% 
  summarise(
    Av_Proportion = mean(Prop_Type, na.rm = TRUE),
    SE_Proportion = sd(Prop_Type,   na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>% 
  # title‐case country, reorder types by their overall mean if you like
  mutate(
    CountryResponsible = str_to_title(CountryResponsible),
    Litter_type       = reorder(Litter_type, -Av_Proportion)
  ) %>% 
  ggplot(data = ., aes(x = Litter_type, y = Av_Proportion)) +
  geom_col(fill = "lightgray", colour = "black") +
  geom_errorbar(aes(ymin = Av_Proportion - SE_Proportion,
                    ymax = Av_Proportion + SE_Proportion), width = 0.2) +
  facet_wrap(~ CountryResponsible, ncol   = 3, scales = "free_y") +
  labs(x = "", y = expression("Single use plastic abundance (mean" %+-% "s.e.)"), title = '', subtitle = "") +
  # theme_classic() +
  # theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  theme_classic(base_size = 16) +
  theme(
    # rotate axis labels, center title, etc.
    axis.text.x     = element_text(angle = 65, hjust = 1),
    plot.title      = element_text(hjust = 0.5),
    strip.background = element_rect(
      fill   = "lightgrey",  # background color
      color  = "black",      # border color
      size   = 0.5           # border thickness
    ),
    strip.text       = element_text(
      colour = "black",      # text color
      face   = "bold"        # make it stand out
    )
  )

```

###### SUP Multivariate Perspective

```{r DataPreparationMultivariateSUP}
# assume your data is in a data.frame called `df`
# and has at least:
#   CountryResponsible | PlasticBags | PlasticBottles | ... | LocalWaste

df_SUP <- GroupB_SingleUsePlastic_data_validated %>% 
  select(CountryResponsible, PlasticBags:Polystyrene)
# %>% 
# #drop_na()
# mutate_all(funs(replace_na(.,0)))

# isolate metadata + community matrix
meta_SUP <- df_SUP %>% select(CountryResponsible)
comm_SUP <- df_SUP %>% select(-CountryResponsible)

# Hellinger‐transform community data (recommended for PCA/cluster/PERMANOVA)
comm_hel_SUP <- decostand(comm, method = "hellinger")

```


In order to see if there are significant differences across country in the community composition, we will perform a PERMANOVA test:

```{r PermanovaTest_SUP, message=FALSE, warning=FALSE}
# 2a) Compute distance and hierarchical clustering (Ward’s D2)
dist_mat_SUP <- vegdist(comm_hel_SUP, method = "euclid")

# 3a) PERMANOVA: test if composition differs by Country
set.seed(42)

perm1_SUP <- adonis2(dist_mat_SUP ~ CountryResponsible, data = meta, 
                 permutations = 999, by = "margin")
print(perm1_SUP)
```

So, we conclude that there are significant differences among countries. The R-Square value (0.098) means that the country "effects" explain 9.8% of the total variability. 90.2% is the remaining (unexplained) variability among countries. ***There is a significant country effect, but is small***.

Or even better:
**There are statistically significant multivariate differences in litter composition among countries, although the effect size is modest (R² = 0.098).**

We can look for **heterocedasticity** among countries

```{r betadisper}
SUP_Disper <- betadisper(dist_mat_SUP, meta_SUP$CountryResponsible)

SUP_Disper
```

```{r}
permutest(SUP_Disper)
```

The **heterocedasticity** is significant, so we must be carefully conclude about the inter-countries differences. There is a Country effect butis small.

If we need to parametrize this model, we should use a Mixed Multivariate Model (necessary?)

```{r mvabund, echo=FALSE, eval=FALSE}
# modelo lineal multivariado generalizado (PERMANOVA Multivariado)
library(mvabund)

mv_data <- mvabund(comm_matrix)

fit <- manyglm(mv_data ~ CountryResponsible,
               family = "negative.binomial")

anova(fit, p.uni = "adjusted")

# Modelo mixto jerárquico bayesiano

brm(
  mvbind(PlasticBags, PlasticBottles, ...) ~ 1 + (1|CountryResponsible),
  family = negbinomial(),
  data = data
)
```


but What are the items that explains these differences? 

The answer will be addressed by performing a **Multivariate Random Forest**.

```{r MultRandForSUP, message=FALSE, warning=FALSE}
# 3b) Multivariate Random Forest:  
#     predict Country from composition → variable importance
rf_SUP <- randomForest(
  x      = comm_hel_SUP,
  y      = as.factor(meta_SUP$CountryResponsible),
  ntree  = 500,
  importance = TRUE
)

print(rf_SUP)
```
The estimated error rate (OOB) is high, so conclusion have to be taken with care. Thus, the more important (differential across countries) litter components are:

```{r RFVarImportSUP, message=FALSE, warning=FALSE}
varImpPlot(rf_SUP, main="Random Forest: Litter‐type Importance for Country Discrimination")
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Random_Forest_Importance_SUP.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
varImpPlot(rf_SUP, main="Random Forest: S.U.P. type Importance for Country Discrimination")
dev.off()
```

This plot of variable importance importance is telling us:

*  A large `MeanDecreaseAccuracy` (**MDA**) means that, when you break the relationship between that predictor and the response, the model’s accuracy suffers a lot → it’s an important predictor for correct country assignments. 
  * **MDA** is generally the more reliable gauge of predictive importance, since it directly measures impact on OOB‐accuracy.

* A large `MeanDecreaseGini` (**MDG**) means that X frequently appears in high‑up the tree splits and yields big purity gains → it’s structurally important to partitioning the data. 
  * **MDG** is faster to compute and tells you which features the trees “prefer” splitting on, but can overweight high–cardinality or continuous variables.
  
Finally, we will perform a Principal Component Analysis (PCA) to see correlation among litter items and *possible* associations with countries

```{r PCASUP, warning=FALSE, message=FALSE}
# ─── 1) PCA (and optional RDA constrained by Country) ───────────────────────────
# 1a) Unconstrained PCA on Hellinger data
# pca1 <- rda(comm_hel)  
# 
# summary(pca1)
# 
# plot(pca1)

# (re)compute PCA with prcomp
pca_prcomp_SUP <- prcomp(comm_hel_SUP, scale. = TRUE)

#summary(pca_prcomp_SUP)
```


```{r PCAplotSUP, warning=FALSE, message=FALSE}
# # 1b) Quick biplot via factoextra
# SUP_PCA_Litter_Country <- 
# fviz_pca_biplot(pca_prcomp_SUP,
#                 geom.ind    = "point",            # show sites as points
#                 col.ind     = meta_SUP$CountryResponsible %>% str_to_title(), 
#                 palette.ind = "jco",              # color‐blind‐friendly palette
#                 addEllipses = FALSE,               # 95% CI per country
#                 label      = "var",               # show arrows for variables
#                 repel      = TRUE
# ) +
#   labs(title = "PCA of Litter Composition (Hellinger‐transformed)",
#        subtitle = "Points = samples colored by Country; arrows = Single use plastic types") +
#   theme_minimal(base_size = 14)
# 
# SUP_PCA_Litter_Country
```


```{r PCAplot2SUP, warning=FALSE, message=FALSE}
# -----------------------------
# Preparar factor y paleta común
# -----------------------------
meta_SUP$Country <- factor(str_to_title(meta_SUP$CountryResponsible))
countries_sup <- levels(meta_SUP$Country)

my_palette_sup <- setNames(
  brewer.pal(length(countries_sup), "Set3"),
  countries_sup
)

# -----------------------------
# Crear gráfico base (p_sup) con puntos pequeños
# -----------------------------
p_sup <- fviz_pca_biplot(
  pca_prcomp_SUP,
  geom.ind    = "point",
  col.ind     = meta_SUP$Country,
  addEllipses = FALSE,
  label       = "var",
  repel       = TRUE,
  pointsize   = 2.5,
  pointshape  = 16,
  alpha.ind   = 0.6
)

# -----------------------------
# Calcular centroides manualmente
# -----------------------------
scores_sup <- as.data.frame(pca_prcomp_SUP$x)
scores_sup$Country <- meta_SUP$Country

centroids_sup <- scores_sup %>%
  group_by(Country) %>%
  summarise(
    PC1 = mean(PC1),
    PC2 = mean(PC2),
    .groups = "drop"
  )

# -----------------------------
# Añadir centroides y unificar escalas
# -----------------------------
SUP_PCA_Litter_Country2 <-
  p_sup +
  geom_point(data = centroids_sup,
             aes(x = PC1, y = PC2, fill = Country),
             shape = 21,
             size = 5,
             color = "black",
             stroke = 1.2) +
  scale_color_manual(values = my_palette_sup, drop = FALSE) +
  scale_fill_manual(values = my_palette_sup, drop = FALSE) +
  labs(title = "PCA of Litter Composition (Hellinger‐transformed)",
       subtitle = "Points = samples colored by Country; arrows = Single use plastic types",
       color = "Country",
       fill  = "Country") +
  theme_minimal(base_size = 14)

SUP_PCA_Litter_Country2
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/PCA_Litter_Country_SUP.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
SUP_PCA_Litter_Country2
dev.off()
```

And the most important Items for the two forst principal components are:

* First PC
```{r PC1VarImportSUP, message=FALSE, warning=FALSE}
# Bar‐plot of top 10 contributors to PC1
fviz_contrib(pca_prcomp_SUP, choice = "var", axes = 1, top = 10) +
  ggtitle("Top 10 variables contributing to PC1")
```

```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Top10_PC1_PCA_Litter_Country_SUP.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
# Bar‐plot of top 10 contributors to PC1
fviz_contrib(pca_prcomp_SUP, choice = "var", axes = 1, top = 10) +
  ggtitle("Top 10 variables contributing to PC1")
dev.off()
```


```{r PC2VarImportSUP, message=FALSE, warning=FALSE}
# Bar‐plot of top 10 contributors to PC2
fviz_contrib(pca_prcomp_SUP, choice = "var", axes = 2, top = 10) +
  ggtitle("Top 10 variables contributing to PC2")
```


```{r, echo=FALSE, eval=FALSE}
# Save the plot
png(filename = "OUTPUT/FIGURES/Top10_PC2_PCA_Litter_Country_SUP.png", width = 30, height = 20, units = "cm", pointsize = 12, bg = "white", res = 150)
# Bar‐plot of top 10 contributors to PC2
fviz_contrib(pca_prcomp_SUP, choice = "var", axes = 2, top = 10) +
  ggtitle("Top 10 variables contributing to PC2")
dev.off()
```


## Floating DATA

### Data loading

```{r FloatingData, message=FALSE, warning=FALSE}
# Cargamos los datos incorporando las modificaciones para trabajar con él
# https://zenodo.org/records/15535434
Floating_data <- read_csv("PlasticPirates_Floating_data/Floating_data_v2.0.0.csv", na = "na") %>%
  select(DatasetID,
    CountryResponsible,
    SamplingYear,
    SamplingAttempted,
    DatasetQualified,
    SamplingDate,
    SamplingCoordinatesLatitude,
    SamplingCoordinatesLongitude,
    SamplingRiverSystem,
    FloatingMacro_AttemptedSampling,
    FloatingMacro_Qualified,
    FloatingMacro_ObservationTimeInMinutes,
    FloatingMacro_RiverWidth,
    FloatingMacro_ObservedWidth,
    FloatingMacro_AmountOfFloatingMacrolitter
  ) %>%
  filter(
    SamplingAttempted == 1 &
    DatasetQualified == 1 &
    FloatingMacro_AttemptedSampling == 1 &
    FloatingMacro_Qualified == 1
  ) %>%
  mutate(across(where(is.character), as.factor),
         CountryResponsible = str_to_title(CountryResponsible))

Floating_data
```

#### Sampling Effor for Floating Data

```{r}
Floating_data %>%
  count(CountryResponsible, SamplingYear) %>%   # contar ambas variables
  ungroup() %>%                                  # importante
  complete(CountryResponsible, SamplingYear, fill = list(n = 0)) %>% 
  ggplot(aes(x = reorder(CountryResponsible, -n),
             y = n,
             fill = factor(SamplingYear))) +
  geom_col(position = "dodge", colour = "black") +
  # Etiquetas encima de cada barra
  geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, # coloca el texto encima
    color = "black", # color negro como quieres 
    fontface = "bold", size = 3
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("2021" = "#E0FFFF",
                               "2022" = "#B8D4F0",
                               "2023" = "#4A90E2",
                               "2024" = "#1E3A8A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "",
       y = "Floating litter sampling (number of samplings)",
       fill = "Sampling Year")
```

### Pre-proccessing

```{r RiversName, message=FALSE, warning=FALSE}
# Loading the rivers' dictionary
Riverbank_dictionary <- read_excel("Riverbank_data_v2.1.xlsx", sheet = "Final vocabulary clean")

Riverbank <-
  Riverbank %>%
  left_join(Riverbank_dictionary, by = "SamplingRiver") %>%
    #select(SamplingRiver, Corrected_river_English, SamplingRiverSystem, River_basin, everything()) %>% 
    mutate(River_basin = str_remove(River_basin, regex("\\s+Basin$", ignore_case = TRUE))) %>% 
  mutate(SamplingRiver = coalesce(Corrected_river_English, SamplingRiver),
         SamplingRiverSystem = River_basin) %>% 
  select(-c("No", "Country", "Corrected_river", "Corrected_river_English", "River_basin", "Count of SamplingRiver"))

```


## RESEARCH QUESTIONS
